---
title: "Short Paper Two (Revision)"
subtitle: "Due on May 13, 2024 by 5pm"
author: Isabelle Gibson
format: 
  pdf:
    fig-pos: "H"
    linestretch: 2
    fontsize: "12pt"
    margin-bottom: "1in"
    margin-top: "1in"
    margin-left: "1in"
    margin-right: "1in"
editor: visual
execute: 
  message: false
  warning: false
  echo: false
---

```{r}
#data 
library(haven)
library(modelsummary)
uscensus <- read_dta("aggcpsgvpt722.dta")

uscensus_shortp2 <- uscensus
```

First, run the following model (you do not need to report the results):

$$
vote = \beta_0+ \beta_1pcthsg + \beta_2south + \beta_3pcthsg*south + \beta_4close  
$$

1.  Write an equation using the estimated coefficient results to show the effect of high school graduation rates on turnout (2 points)

$$
vote = \beta_2 + \beta_3*south
$$

$$
vote = -22.204 + 0.234*south
$$

2.  Write an equation using the estimated coefficient results to show the effect of south on turnout (2 points)

$$
vote = \beta_1 + \beta_3*pcthsg
$$

$$
vote = 0.015 + 0.234*pcthsg
$$

3.  Can $\beta_2$ be interpreted on its own? If so, what is it? If not, why not? Answer in no more than 1 sentence. (2 points)

-   **ANSWER:** Yes, b2 can be interpreted on its own because it is the effect of high school graduation rates on the non-south and south variable.

4.  Is the effect of high school graduation rates on turnout statistically significant? How do you know (answer in no more than 1 sentence and report the relevant test statistics) (2 points)

$H_0$ = High school graduation rates and the interaction with southern state status have no effect on voter turnout

-   **ANSWER:** Yes it is statistically significant, when examining a linear hypothesis test in Table 2, the F-statistic of 7.0388 with a highly significant p-value of 0.001006 clearly rejects the null hypothesis confirming significant effects in both high school graduation rates alone **and** in combination with being in a southern state on the variation in voter turnout.

5.  Is the effect of high school graduation rates on turnout substantively significant in the south? What about the non-south? Answer in no more than 4 sentences total? (5 points)

$$
vote = \beta_0+ \beta_1pcthsg + (-22.204)south + \beta_3pcthsg*south + \beta_4close 
$$

-   **SOUTH:** The coefficient on Southern states in Model 2 is -22.204 (Table 1) indicates that on average southern states voter turnout decreases by 22.204 percentage points, holding all else constant.

    -   This is substantively significant effect because 22.204 percentage points in an election can change election outcomes due soley to the geographic location of being in the south.

-   **NON-SOUTH:** The Non-southern states in Model 2 coefficient (0.015) indicates that on average non-southern states voter turnout increases by 0.015 percentage points, holding all else constant.

    -   This suggests that any change in high school graduation rates is not substantively significant, an average 0.015 percentage point increase in voter turnout in non-southern states wont likely effect electoral outcomes

#### Appendix

**Table 1**

```{r}
#run the model 

reg_model <- lm(vote ~ close + pcthsg + south, data = uscensus_shortp2)
south_HS_model <- lm(vote ~ close + pcthsg + south + pcthsg:south, data = uscensus_shortp2)

newmodel <- list(
  "Model 1." = reg_model,
  "Model 2" = south_HS_model
)

modelsummary(newmodel,
             position = "H",
             title = "Linear regression model of Voter turnout",
             coef_rename = c(south = "Southern States",
                             pcthsg = "Percentage of HS Graduates",
                             close = "Voter Registration Closing Date",
                             "pcthsg:south" = "Percentage of HS Graduates - Southern State"),
          conf_level = .95)
```

**Table 2**

```{r}
#linear hypothesis test pcthsg 

library(broom)
library(car)
library(kableExtra)
library(dplyr)

# Perform linear hypothesis test
hypo1 <- linearHypothesis(south_HS_model, c("pcthsg = 0", "pcthsg:south = 0"))

# Format p-values
hypo1$`Pr(>F)` <- as.character(format(hypo1$`Pr(>F)`, scientific = FALSE))
hypo1$`Pr(>F)`[hypo1$`Pr(>F)` < 0.001] <- paste0(hypo1$`Pr(>F)`[hypo1$`Pr(>F)` < 0.001], "***")
hypo1$`Pr(>F)`[hypo1$`Pr(>F)` < 0.01 & hypo1$`Pr(>F)` >= 0.001] <- paste0(hypo1$`Pr(>F)`[hypo1$`Pr(>F)` < 0.01 & hypo1$`Pr(>F)` >= 0.001], "**")
hypo1$`Pr(>F)`[hypo1$`Pr(>F)` < 0.05 & hypo1$`Pr(>F)` >= 0.01] <- paste0(hypo1$`Pr(>F)`[hypo1$`Pr(>F)` < 0.05 & hypo1$`Pr(>F)` >= 0.01], "*")

# Create LaTeX table
kable_out <- kable(hypo1, format = "latex", booktabs = TRUE, caption = "The Linear Hypothesis test of High school Graduation Rates")

# Add footnotes
kable_out <- footnote(kable_out, general = c("*** p < 0.001", "** p < 0.01", "* p < 0.05"),
                      threeparttable = TRUE)

kable_out
```
