---
title: "Short Paper Two"
author: Isabelle Gibson
format: 
  pdf:
    fig-pos: "H"
    linestretch: 2
    fontsize: "12pt"
    margin-bottom: "1in"
    margin-top: "1in"
    margin-left: "1in"
    margin-right: "1in"
editor: visual
execute: 
  message: false
  warning: false
  echo: false
---

```{r}
library(foreign)
library(dplyr)
library(ggplot2)
library(haven)
library(modelsummary)
library(broom)
library(sandwich)
library(car)
library(marginaleffects)
library(kableExtra)
library(car)
library(knitr)
```

```{r}
#data
uscensus <- read_dta("aggcpsgvpt722.dta")

uscensus_shortp2 <- uscensus
```

***Expectations***

Reviewer noted that I need to evaluate the expectation that the effects of high school graduation rates, and that a state being southern or not are not conditional on one another. To test this relationship, I provide scatter plots with fitted lines of the relationship between high school graduation rates and turnout in the south and non-south, without controlling for other variables. Then, I ran a model that extends to includes an interaction terms between high school graduation rate and south variables. I expect that the negative effect of the south becomes less strong as high school graduation rates increase (i.e. the effect of the south is less strongly negative at higher levels of high school graduation rates).

1.  Provide scatter plots with fitted regression lines of the relationship between high school graduation rates and turnout in the south and non-south, without controlling for any other variables. Does it appear from this descriptive analysis that the effect of high school graduation rates is different in the south and non-south?

-   The scatter-plot in Figure 1 (see appendix) highlights the positive association between the percentage of high school graduation rates and voter turnout. The scatter-plot presents that as high school graduation rates increase, voter turnout increases in the south. In other words, the **negative effect of being in the South** becomes **weaker** as the percentage of high school graduates rises. At higher levels of high school graduations rates, the negative impact of being in the south on voter turnout is less pronounced. This provides support for the reviews hypothesis that negative effect of the south becomes less strong as high school graduation rates increase (i.e. the effect of the south is less strongly negative at higher levels of high school graduation rates).

Table 1 (see appendix) presents the original model (Model 1):

> $$
> Vote = \beta_0\ +close \times\ \beta_1\ + pcthsg \times\ \beta_2\ + south \times\ \beta_3\ + e 
> $$
>
> And the new interaction model (Model 2):
>
> $$
> Vote = \beta_0+\beta_1(close)+\beta_2(pcthsg)+\beta_3(south)*(pcthsg)+\epsilon_i 
> $$

2.  Does the evidence support my hypothesis? Be sure to discuss both substantive and statistical significance and include a graph of the effect of the south on turnout across high school graduation rates with confidence intervals around the effect

-   The **interaction term** between **percentage of** **high school graduation** and **Southern states** in my new regression model has a positive coefficient of **0.234** and is statistically significant at p \< 0.01. The positive coefficient (0.234) on the interaction term suggests that the negative effect associated with being in a Southern state on voter turnout decreases as the high school graduation rates increase. This means that higher percentages of high school graduation in a state can mitigate some of the challenges of voter turnout that might be more prevalent in Southern states. These findings support the hypothesis that the negative effect of the south becomes less strong as high school graduation rates increase.

-   Figure 2 (see appendix) provides a graph of the effect of south on turnout across high school graduation rates with confidence intervals around the effect. It gives as a visual representation of the interaction between south and high school graduation rates on voter turnout. The steeper slope for Southern states (red line) compared to non-southern states (blue line) suggests that high school graduation rates have a more significant effect on increasing voter turnout in the south.

3.  What is the substantive meaning of the coefficient on the interaction term between high school graduation rates and south?

-   The interaction coefficient of **Percentage of HS Graduates - Southern State is** .234. It implies that for each additional percentage point increase in high school graduation rates, the negative impact of being a southern state on voter turn out is reduced by about .234 **percentage points** **on average, holding all else constant.** At a first glance, this may seem small and not substantively significant, but this effect can accumulate into a significant reduction of the disparity.

4.  **Overall, based on the new model, is the effect of high school graduation rates on turnout statistically significant? Please report the relevant test statistics.**

-   Table 1 (see appendix) shows the new model and indicates that the influence of high school graduation rates on voter turnout has a coefficient of 0.015, which is not statistically significant at p \< 0.05. However, the interaction between high school graduation rates and Southern States is significant, with a coefficient of 0.234. This suggests that the impact of high school graduation rates on voter turnout differs depending on whether a state is in the South. The t-stat is .376, it is not significantly different from 0, indicating that there is very weak evidence against the null hypothesis (H0= there is no effect of the pcthsg on voter turnout).

-   In Table 2, we see an F-statistic of 7.0388 with a corresponding p-value of 0.001006, indicating high significance. Since this p-value is much less than 0.05, we reject the null hypothesis. This means that high school graduation rates indeed have a significant effect on voter turnout, and there's also a significant interaction between high school graduation rates and Southern status in explaining voter turnout variability.

-   The data suggests that high school graduation rates play a significant role in voter turnout, especially when we consider their interaction with Southern status.

-   Although the primary effect is not significant, the significant interaction term indicates that the relationship between high school graduation rates and voter turnout is region-dependent, specifically emphasizing a conditional effect associated with being in the South.

5.  Is the effect of high school graduation rates on turnout substantively significant in the south? What about the non-south?

-   The coefficient on Southern states in Model 2 is -22.204 indicating that **on average** southern states have a significantly lower voter turnout, **holding all else constant,** compared to non-southern states. **This is a large substantive effect**. This is substantively significant because it suggests that a targeted approach for policy-makers, educational policies might not only have educational benefits but also civic and social benefits in terms of increased participation on voting.

-   The Non-southern states in Model 2 coefficient (0.015) is not statistically significant. This suggests that any change in high school graduation rates **does not have a significant direct effect on turnout in non-Southern states**. Different factors may play a more significant role in voter turnout in non-southern states.

6.  Overall, based on the new model, is the effect of south on turnout statistically significant? **Please report the relevant test statistics**

-   The new model shows a negative impact of being a Southern State on voter turnout and is statistically significant. in Table 1: Model 2 (see appendix), the coefficient for Southern States is -22.204, indicating a p-value of less than 0.001. The negative coefficient indicates that being a Southern state is associated with lower voter turnout compared to non-Southern states. This finding remains significant even after adjusting for high school graduation rates and their interaction with being in the South, underscoring the impact of a state's geographic (south or non-south) location on voter turnout.

-   The t-stat is -3.876 (Table 1, see appendix) indicates that the coefficient is significantly different from 0, the negative t-stat indicates that the coefficient is less than 0, this suggests a negative effect of being a southern state on voter turnout, and we observe this relationship.

-   When examining the F-statistic of 28.034, presented in Table 3, an extremely small p-value is evident. This p-value indicates strong evidence against the null hypothesis, presenting a substantial effect of the south on turnout, alongside the existence of an interaction effect between high school graduation rates and southern status.

-   The linear hypothesis test the findings suggest that whether Southern or non-Southern, along with the interaction between high school graduation rates and southern status, significantly influence voter turnout within the new model. The variables denoting Southern status and its interaction with high school graduation rates are influential factors in clarifying voter turnout patterns within the model.

7.  Finally, please explain in a few sentences why you will or will not include the new model in the paper.

-   I will include the new model in the paper, because it provides a better understanding of the dynamics between percentage of high school graduation rates and being in a Southern states on voter turnout. The significant interaction term confirms the hypothesis that the negative effect of being in the South on voting rates diminishes as high school graduation rates increase. This suggests that high school graduation rates in Southern states can mitigate some of the negative factors influencing voting rates in these states.

***Appendix***

***Figure 1: The relationship between high school graduation rates and voter turnout in the south and non-south regions using scatter plots with fitted regression lines***

```{r}
#HS grad % and voter turnout in South regions 
# Scatter plot for the South region (coded as 1)

ggplot(filter(uscensus_shortp2), aes(x = pcthsg, y = vote, color = factor(south))) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Figure 1 Scatterplot: Voter Turnout Rate vs. % of HS Graduates",
       x = "(%) of High School Graduates",
       y = "Voter Turnout Rate") +
  theme_minimal() +
  labs(
    x = "% of HS Graduates",
    y = "Voter Turnout"
  ) +
  scale_color_manual(values = c("blue", "red"),labels = c("Non-Southern States", "Southern States")
                     ) +
  guides(color = guide_legend(title = NULL))
```

***Table 1: Original Multivariate Regression Model and New Interaction model with High School Graduation Rates and Southern State variable Interaction Terms***

```{r}
reg_model <- lm(vote ~ close + pcthsg + south, data = uscensus_shortp2)
south_HS_model <- lm(vote ~ close + pcthsg + south + pcthsg:south, data = uscensus_shortp2)

newmodel <- list(
  "Model 1." = reg_model,
  "Model 2" = south_HS_model
)

modelsummary(newmodel,
             position = "H",
             title = "Original Model without interaction terms & New model with Interaction terms",  stars = T, statistic = c("p.value", "t = {statistic}"),
             coef_rename = c(south = "Southern States",
                             pcthsg = "Percentage of HS Graduates",
                             close = "Voter Registration Closing Date",
                             "pcthsg:south" = "Percentage of HS Graduates - Southern State"),
          conf_level = .95)
```

***Figure 2. Plotted predictions of the effect of south on voter turnout across high school graduation rates with confidence intervals around the effect***

```{r}
#plot prediction 
plot_predictions(south_HS_model, condition = c("pcthsg", "south")) + 
  theme_minimal() + 
  labs(colour = "south (1) non-south (0)" ,
       x = "High school Graduation rates",
       y = "Voter Turnout",
       title = "Effect of South on Voter turnout across HS Graduation Rates") +
  scale_color_manual(values = c("blue", "red"),labels = c("Non-Southern States", "Southern States")) +
                       scale_fill_manual(values = c("blue", "red")) + guides(fill = FALSE)

```

```{r}
#linear hypothesis test pcthsg 

library(broom)
library(car)
library(kableExtra)

# Perform linear hypothesis test
hypo1 <- linearHypothesis(south_HS_model, c("pcthsg = 0", "pcthsg:south = 0"))

# Format p-values
hypo1$`Pr(>F)` <- as.character(format(hypo1$`Pr(>F)`, scientific = FALSE))
hypo1$`Pr(>F)`[hypo1$`Pr(>F)` < 0.001] <- paste0(hypo1$`Pr(>F)`[hypo1$`Pr(>F)` < 0.001], "***")
hypo1$`Pr(>F)`[hypo1$`Pr(>F)` < 0.01 & hypo1$`Pr(>F)` >= 0.001] <- paste0(hypo1$`Pr(>F)`[hypo1$`Pr(>F)` < 0.01 & hypo1$`Pr(>F)` >= 0.001], "**")
hypo1$`Pr(>F)`[hypo1$`Pr(>F)` < 0.05 & hypo1$`Pr(>F)` >= 0.01] <- paste0(hypo1$`Pr(>F)`[hypo1$`Pr(>F)` < 0.05 & hypo1$`Pr(>F)` >= 0.01], "*")

# Create LaTeX table
kable_out <- kable(hypo1, format = "latex", booktabs = TRUE, caption = "The effect of High School Graduation rates on Turnout") %>%
  add_header_above(c("Linear Hypothesis Test" = 6))

# Add footnotes
kable_out <- footnote(kable_out, general = c("*** p < 0.001", "** p < 0.01", "* p < 0.05"),
                      threeparttable = TRUE)

kable_out


```

```{r}
#south 
library(broom)
library(car)
library(kableExtra)

# Perform linear hypothesis test
hypo2 <- linearHypothesis(south_HS_model, c("south = 0", "pcthsg:south = 0"))

# Format p-values
hypo2$`Pr(>F)` <- as.character(format(hypo2$`Pr(>F)`, scientific = FALSE))
hypo2$`Pr(>F)`[hypo2$`Pr(>F)` < 0.001] <- paste0(hypo2$`Pr(>F)`[hypo2$`Pr(>F)` < 0.001], "***")
hypo2$`Pr(>F)`[hypo2$`Pr(>F)` < 0.01 & hypo2$`Pr(>F)` >= 0.001] <- paste0(hypo2$`Pr(>F)`[hypo2$`Pr(>F)` < 0.01 & hypo2$`Pr(>F)` >= 0.001], "**")
hypo2$`Pr(>F)`[hypo2$`Pr(>F)` < 0.05 & hypo2$`Pr(>F)` >= 0.01] <- paste0(hypo2$`Pr(>F)`[hypo2$`Pr(>F)` < 0.05 & hypo2$`Pr(>F)` >= 0.01], "*")

# Create LaTeX table
kable_out2 <- kable(hypo2, format = "latex", booktabs = TRUE, caption = "The effect of Southern States on Turnout") %>%
  add_header_above(c("Linear Hypothesis Test" = 6))

# Add footnotes
kable_out2 <- footnote(kable_out2, general = c("*** p < 0.001", "** p < 0.01", "* p < 0.05"))

kable_out2
```
